The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-02-15 script/en/chapters/ellipticcurves.tex
Corrected typos and overfull hbox; congruent index (no redundancy in wording; adjectives in lowercase in German; all normal words in lowercase in English)

\newpage
\hypertarget{ellcurve}{}
\chapter{Elliptic Curves}
\label{Chapter_EllipticCurves}
\index{Elliptic curves}
(Filipovics B. / B\"uger M. / Esslinger B. / Oyono R., April 2000,
Updates: Dec. 2001, June 2002, Mar. 2003, Nov. 2009, Aug. 2013)

\section{Elliptic curve cryptography -- a high-performance substitute for RSA?}
\index{Performance}
\label{ECAlternative}

@@

Anyway that's enough about the role of elliptic curves in pure mathematics. 
In 1985 Neal Koblitz\index{Koblitz, Neal} and Victor Miller\index{Miller, Victor}
independently suggested using elliptic curves in cryptography. Elliptic 
curves have thus also found a concrete practical application. 
Another interesting area of application for elliptic curves is for
factorizing whole numbers \index{Factorization} (the RSA cryptographic system
is based on the \index{Complexity} difficulty/complexity of finding prime
factors of an extremely large number;  compare section \ref{SecurityRSA}.).
In this area, procedures based on elliptic curves have been investigated and
used since 1987 (compare section \ref{ECC-Factorization}). \\
There are also prime number tests\index{Prime number!test} based on elliptic
curves.
% (s. Anmerkung U. Kuehn)

Elliptic curves are used differently in the various areas. Encryption
procedures based on elliptic curves are based on the difficulty of a problem
known as elliptic curve discrete logarithm\index{Logarithm problem!discrete}.
The factorization of whole numbers uses the fact that a large number of
elliptic curves can be generated for a natural composite number $n$ with
several prime factors; however, these curves are not then groups for composite
$n$. More information about this can be found under the chapter \ref{ECC-Factorization}.

@@

This section provides information about \index{Group} {\em groups} and
\index{field} {\em fields}.

@@

\paragraph*{Cyclic groups}\index{Group!cyclic}
Cyclic groups\footnote{Cyclic groups can be in general also endless like the additive group of the integer numbers. We consider here only finite cyclic groups.} are those groups $G'$ that possess an element $g$
from which the group operation can be used to generate all other
elements in the group. This means that for each element $a$ in
$G'$ there exists a positive whole number $i$ such that if $g$ is
subject to the operation $i$ times (i.e. ``$g\cdot i$''),
$g+g+\cdots+g = a$ (additive group) or $g^i = g\cdot g \cdots g = a$
(multiplicative group). The element $g$ is the {\em generator} of
the cyclic group --- each element in $G'$ can be generated using
$g$ and the operation.

@@

In cryptography elliptic curve are a useful tool. Such curves are described by some equation. A detailed analysis has shown that curves of the form\footnote{This curve is given by the zeros of a {\it polynomial}\index{Polynomial} $F$ of degree three in three variables. In general, expressions of the form
$P=\sum_{i_1,\dots,i_n\in\N_0} a_{i_1\dots i_n} x_1^{i_1}\dots x_n^{i_n}$ with coefficients $a_{i_1\dots i_n}\in K$ are called polynomials in $n$ variables $x_1,\dots,x_n$ with underlying field $K$, if ${\rm deg\,} P:=\max\{i_1+\dots +i_n: a_{i_1\dots i_n}\ne 0\}$ is finite, i.e. the sum has only finitely many non-zero terms (monomials). The sum of the exponents of the variables of each term of the sum is at most $3$, at least one term of the sum has a single variable with $3$ as value of the according exponent.}
\begin{equation}
 F(x_1,x_2,x_3)=-x_1^3+x_2^2x_3+a_1x_1x_2x_3-a_2x_1^2x_3+a_3x_2x_3^2-a_4x_1x_3^2-a_6x_3^3=0,
\label{eccbasisgleichung}
\end{equation}
are especially useful. The variables $x_1,x_2,x_3$ and parameters $a_1,\dots,a_4,a_6$ are elements of a given field $K$, which has certain properties that are make it useful from the cryptographic point of view. The underlying field $K$ might be the well known field of real numbers or some finite field (see last section).
In order to obtain a cure that is useful for cryptography, the parameters have to be chosen in a way that the following conditions hold
$$ \frac{\partial F}{\partial x_1}\ne 0, \quad \frac{\partial F}{\partial x_2}\ne 0, \quad
\frac{\partial F}{\partial x_3}\ne 0 .
$$
We identify points on the curve that can be derived from each over by multiplying each component with some scalar. This makes sense since $(x_1,x_2,x_3)$ solves (\ref{eccbasisgleichung}) if and only if $\alpha (x_1,x_2,x_3)$ ($\alpha\ne 0$) does. Formally, this means that we consider classes of equivalent points instead of single points, where points are called equivalent if one is the scalar multiple of the other one.
\\ If we put $x_3=0$ in the basic equation (\ref{eccbasisgleichung}), then this equation collapses to $-x_1^3=0$, leading to $x_1=0$. Thus, the equivalence class which includes the element $(0,1,0)$ is the only one that contains a point with $x_3=0$. For all points on the curve that are not equivalent to $(0,1,0)$, we may apply the following transformation
$$ K\times K\times (K\setminus\{0\})\ni (x_1,x_2,x_3) \mapsto (x,y):=\left( \frac{x_1}{x_3}, \frac{x_2}{x_3}\right) \in K\times K \, ,
$$
which reduces the number of variables to two instead of three. We note that the
basic equation (\ref{eccbasisgleichung}) $F(x_1,x_2,x_3)=0$  was chosen in a
way that this transformation leads to the famous so-called
Weierstrass-Equation\footnote{Karl Weierstrass\index{Weierstrass, Karl}, 31.10.1815$-$19.12.1897, German
mathematician, famous for his rigorous formal approach to mathematics.} holds
\begin{equation}
 y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6 \, .
\label{ell}
\end{equation}
Since all but one point (i.e. equivalence class) of the elliptic curve can be described using equation (\ref{ell}), this equation is often called the elliptic equation, and its solutions written as
$$ {\bf E} = \left\{(x,y)\in K\times K \, |\, y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6  \right\} \cup \{{\cal O} \}.
$$
Here, ${\cal O}$ represents the point $(0,1,0)$ that is loosely speaking mapped to infinity by the transformation (division by $x_3$) that reduces the three variables to two.

@@

\begin{sloppypar}
  The {\em elliptic curve discrete logarithm
    problem}\index{Problem of discrete logarithm} (ECDLP) \index{ECDLP} is the basis for elliptic curve cryptography. Based on this problem, there are different signature schemes. In order to apply one of these, we need:
\end{sloppypar}
\begin{itemize}
    \item An elliptic curve {\bf E} with an underlying field $GF(p^n)$.
    \item A prime $q\ne p$ and a point $G$ on the elliptic curve ${\bf E}$ with order $q$. This means that $qG={\cal O}$ and $rG\ne {\cal O}$ for all $r\in \{1,2,\dots,q-1\}$. Thus $q$ is a factor of the group order (i.e. the number of elements) $\#{\bf E}$ of $E$. Since $q$ is prime, $G$ generates a cyclic sub-group of ${\bf E}$ of order $q$.
\end{itemize}
The parameters mentioned are often called \index{Domain parameter}
{\em Domain} parameter. They describe the elliptic curve ${\bf E}$ and 
the cyclic sub-group of ${\bf E}$ on which the signature scheme is based.

@@

There are factorization%
\footnote{Especially John M. Pollard \index{Pollard, John M.} was involved
in the development of many different factorization algorithms; also at
factorization with ECC he was one of the leading heads. As an employee
of British Telekom he never published much. At the RSA data Security Conference
in 1999 he was awarded for his ``outstanding contributions in mathematics'':
\url{http://www.eff.org/Privacy/Crypto_misc/DESCracker/HTML/19990118_rsa_awards.html}.\\
In 1987 H.W. Lenstra published an often used factorization algorithm,
based on elliptic curves (see \cite{ec:Lenstra1987}).
}
algorithms based on elliptic curves%
\footnote{The biggest compound numbers currently 
factorized\index{Factorization!factoring records} with elliptic curves have
about 80 decimal digits:\\
\url{http://www.loria.fr/~zimmerma/records/top50.html}.\\
See also the web link about the \hyperlink{Lenstra2}{ECMNET project}\index{ECMNET} at the end of this chapter.
}. 
More precisely, these procedures exploit the fact that elliptic curves can
be defined over ${\mathbb Z}_n$ ($n$ composite number). Elliptic curves 
over ${\mathbb Z}_n$ do not form a group, because not every point on such 
an elliptic curve has an inverse point. This is connected with the fact 
that - if $n$ is a composite
number - there exist elements in ${\mathbb Z}_n$ that do not have an inverse
with respect to multiplication mod $n$. In order to add two points on an
elliptic curve over ${\mathbb Z}_n$, we can calculate in the same way as on
elliptic curves over ${\mathbb Z}_p$. Addition of two points (on an elliptic
curve over ${\mathbb Z}_n$), however, fails if and only if a factor of $n$ has
been found. The reason for this is that the procedure for adding points on
elliptic curves gives elements in ${\mathbb Z}_n$ and calculates the inverse
elements for these (with respect to multiplication mod $n$) in ${\mathbb Z}_n$.
The extended \index{Euclidean algorithm} Euclidean algorithm is used here. If
the addition of two points (that lie of an elliptic curve over ${\mathbb Z}_n$)
gives an element in ${\mathbb Z}_n$ that does not have an inverse element in
${\mathbb Z}_n$, then the extended Euclidean algorithm delivers a genuine factor
of $n$.

@@

% -----------------------------------------------------------------------------
\subsection{SageMath}
\label{ec:Sage_Massierer}
\index{SageMath}
\index{SageMath!Code examples}

@@

The ECC notebook\index{Elliptic curves!ECC notebook}\index{Massierer, Maike} of
Massierer\footnote{%
Instructions to use an interactive SageMath notebook\index{SageMath!Instructions interactive notebook}:\\

@@

% -----------------------------------------------------------------------------
\newpage
\section{Patent aspects}
\index{Patent}

>>>>

\newpage
\hypertarget{ellcurve}{}
\chapter{Elliptic Curves}
\label{Chapter_EllipticCurves}
\index{elliptic curve}
(Filipovics B. / B\"uger M. / Esslinger B. / Oyono R., April 2000,
Updates: Dec. 2001, June 2002, Mar. 2003, Nov. 2009, Aug. 2013)

\section{Elliptic curve cryptography -- a high-performance substitute for RSA?}
\index{performance}
\label{ECAlternative}

@@

Anyway that's enough about the role of elliptic curves in pure mathematics. 
In 1985 Neal Koblitz\index{Koblitz, Neal} and Victor Miller\index{Miller, Victor}
independently suggested using elliptic curves in cryptography. Elliptic 
curves have thus also found a concrete practical application. 
Another interesting area of application for elliptic curves is for
factorizing whole numbers \index{factorization} (the RSA cryptographic system
is based on the \index{complexity} difficulty/complexity of finding prime
factors of an extremely large number;  compare section \ref{SecurityRSA}.).
In this area, procedures based on elliptic curves have been investigated and
used since 1987 (compare section \ref{ECC-Factorization}). \\
There are also prime number tests\index{prime number!test} based on elliptic
curves.
% (s. Anmerkung U. Kuehn)

Elliptic curves are used differently in the various areas. Encryption
procedures based on elliptic curves are based on the difficulty of a problem
known as elliptic curve discrete logarithm\index{logarithm problem!discrete}.
The factorization of whole numbers uses the fact that a large number of
elliptic curves can be generated for a natural composite number $n$ with
several prime factors; however, these curves are not then groups for composite
$n$. More information about this can be found under the chapter \ref{ECC-Factorization}.

@@

This section provides information about \index{group} {\em groups} and
\index{field} {\em fields}.

@@

\paragraph*{Cyclic groups}\index{group!cyclic}
Cyclic groups\footnote{Cyclic groups can be in general also endless like the additive group of the integer numbers. We consider here only finite cyclic groups.} are those groups $G'$ that possess an element $g$
from which the group operation can be used to generate all other
elements in the group. This means that for each element $a$ in
$G'$ there exists a positive whole number $i$ such that if $g$ is
subject to the operation $i$ times (i.e. ``$g\cdot i$''),
$g+g+\cdots+g = a$ (additive group) or $g^i = g\cdot g \cdots g = a$
(multiplicative group). The element $g$ is the {\em generator} of
the cyclic group --- each element in $G'$ can be generated using
$g$ and the operation.

@@

In cryptography elliptic curve are a useful tool. Such curves are described by some equation. A detailed analysis has shown that curves of the form\footnote{This curve is given by the zeros of a {\it polynomial}\index{polynomial} $F$ of degree three in three variables. In general, expressions of the form
$P=\sum_{i_1,\dots,i_n\in\N_0} a_{i_1\dots i_n} x_1^{i_1}\dots x_n^{i_n}$ with coefficients $a_{i_1\dots i_n}\in K$ are called polynomials in $n$ variables $x_1,\dots,x_n$ with underlying field $K$, if ${\rm deg\,} P:=\max\{i_1+\dots +i_n: a_{i_1\dots i_n}\ne 0\}$ is finite, i.e. the sum has only finitely many non-zero terms (monomials). The sum of the exponents of the variables of each term of the sum is at most $3$, at least one term of the sum has a single variable with $3$ as value of the according exponent.}
\begin{equation}
 F(x_1,x_2,x_3)=-x_1^3+x_2^2x_3+a_1x_1x_2x_3-a_2x_1^2x_3+a_3x_2x_3^2-a_4x_1x_3^2-a_6x_3^3=0,
\label{eccbasisgleichung}
\end{equation}
are especially useful. The variables $x_1,x_2,x_3$ and parameters $a_1,\dots,a_4,a_6$ are elements of a given field $K$, which has certain properties that are make it useful from the cryptographic point of view. The underlying field $K$ might be the well known field of real numbers or some finite field (see last section).
In order to obtain a cure that is useful for cryptography, the parameters have to be chosen in a way that the following conditions hold
$$ \frac{\partial F}{\partial x_1}\ne 0, \quad \frac{\partial F}{\partial x_2}\ne 0, \quad
\frac{\partial F}{\partial x_3}\ne 0 .
$$
We identify points on the curve that can be derived from each over by multiplying each component with some scalar. This makes sense since $(x_1,x_2,x_3)$ solves (\ref{eccbasisgleichung}) if and only if $\alpha (x_1,x_2,x_3)$ ($\alpha\ne 0$) does. Formally, this means that we consider classes of equivalent points instead of single points, where points are called equivalent if one is the scalar multiple of the other one.
\\ If we put $x_3=0$ in the basic equation (\ref{eccbasisgleichung}), then this equation collapses to $-x_1^3=0$, leading to $x_1=0$. Thus, the equivalence class which includes the element $(0,1,0)$ is the only one that contains a point with $x_3=0$. For all points on the curve that are not equivalent to $(0,1,0)$, we may apply the following transformation
$$ K\times K\times (K\setminus\{0\})\ni (x_1,x_2,x_3) \mapsto (x,y):=\left( \frac{x_1}{x_3}, \frac{x_2}{x_3}\right) \in K\times K \, ,
$$
which reduces the number of variables to two instead of three. We note that the
basic equation (\ref{eccbasisgleichung}) $F(x_1,x_2,x_3)=0$  was chosen in a
way that this transformation leads to the famous so-called
Weierstrass-Equation\footnote{Karl Weierstrass\index{Weierstrass, Karl}, 31.10.1815$-$19.12.1897, German
mathematician, famous for his rigorous formal approach to mathematics.} holds
\begin{equation}
 y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6 \, .
\label{ell}
\end{equation}
Since all but one point (i.e. equivalence class) of the elliptic curve can be described using equation (\ref{ell}), this equation is often called the elliptic equation, and its solutions written as
$$ {\bf E} = \left\{(x,y)\in K\times K \, |\, y^2+a_1xy+a_3y = x^3+a_2x^2+a_4x+a_6  \right\} \cup \{{\cal O} \}.
$$
Here, ${\cal O}$ represents the point $(0,1,0)$ that is loosely speaking mapped to infinity by the transformation (division by $x_3$) that reduces the three variables to two.

@@

\begin{sloppypar}
  The {\em elliptic curve discrete logarithm
    problem}\index{problem of discrete logarithm} (ECDLP) \index{ECDLP} is the basis for elliptic curve cryptography. Based on this problem, there are different signature schemes. In order to apply one of these, we need:
\end{sloppypar}
\begin{itemize}
    \item An elliptic curve {\bf E} with an underlying field $GF(p^n)$.
    \item A prime $q\ne p$ and a point $G$ on the elliptic curve ${\bf E}$ with order $q$. This means that $qG={\cal O}$ and $rG\ne {\cal O}$ for all $r\in \{1,2,\dots,q-1\}$. Thus $q$ is a factor of the group order (i.e. the number of elements) $\#{\bf E}$ of $E$. Since $q$ is prime, $G$ generates a cyclic sub-group of ${\bf E}$ of order $q$.
\end{itemize}
The parameters mentioned are often called \index{domain parameter}
{\em Domain} parameter. They describe the elliptic curve ${\bf E}$ and 
the cyclic sub-group of ${\bf E}$ on which the signature scheme is based.

@@

There are factorization%
\footnote{Especially John M. Pollard \index{Pollard, John M.} was involved
in the development of many different factorization algorithms; also at
factorization with ECC he was one of the leading heads. As an employee
of British Telekom he never published much. At the RSA data Security Conference
in 1999 he was awarded for his ``outstanding contributions in mathematics'':
\url{http://www.eff.org/Privacy/Crypto_misc/DESCracker/HTML/19990118_rsa_awards.html}.\\
In 1987 H.W. Lenstra published an often used factorization algorithm,
based on elliptic curves (see \cite{ec:Lenstra1987}).
}
algorithms based on elliptic curves%
\footnote{The biggest compound numbers currently 
factorized\index{factorization!factoring records} with elliptic curves have
about 80 decimal digits:\\
\url{http://www.loria.fr/~zimmerma/records/top50.html}.\\
See also the web link about the \hyperlink{Lenstra2}{ECMNET project}\index{ECMNET} at the end of this chapter.
}. 
More precisely, these procedures exploit the fact that elliptic curves can
be defined over ${\mathbb Z}_n$ ($n$ composite number). Elliptic curves 
over ${\mathbb Z}_n$ do not form a group, because not every point on such 
an elliptic curve has an inverse point. This is connected with the fact 
that - if $n$ is a composite
number - there exist elements in ${\mathbb Z}_n$ that do not have an inverse
with respect to multiplication mod $n$. In order to add two points on an
elliptic curve over ${\mathbb Z}_n$, we can calculate in the same way as on
elliptic curves over ${\mathbb Z}_p$. Addition of two points (on an elliptic
curve over ${\mathbb Z}_n$), however, fails if and only if a factor of $n$ has
been found. The reason for this is that the procedure for adding points on
elliptic curves gives elements in ${\mathbb Z}_n$ and calculates the inverse
elements for these (with respect to multiplication mod $n$) in ${\mathbb Z}_n$.
The extended \index{Euclidean algorithm} Euclidean algorithm is used here. If
the addition of two points (that lie of an elliptic curve over ${\mathbb Z}_n$)
gives an element in ${\mathbb Z}_n$ that does not have an inverse element in
${\mathbb Z}_n$, then the extended Euclidean algorithm delivers a genuine factor
of $n$.

@@

% -----------------------------------------------------------------------------
\subsection{SageMath}
\label{ec:Sage_Massierer}
\index{SageMath}
\index{SageMath!code examples}

@@

The ECC notebook\index{elliptic curve!ECC notebook}\index{Massierer, Maike} of
Massierer\footnote{%
Instructions to use an interactive SageMath notebook\index{SageMath!instructions interactive notebook}:\\

@@

% -----------------------------------------------------------------------------
\newpage
\section{Patent aspects}
\index{patent}

### 2016-02-15 script/en/chapters/menus.tex
First version of CT Book ed 12. Most important point is the newly includes bitciphers chapter. Further things: Tried to reduce latex warnings amap. Changed wording from Sage to SageMath where visible...
Corrected typos and overfull hbox; congruent index (no redundancy in wording; adjectives in lowercase in German; all normal words in lowercase in English)

Which of the menu items in CrypTool 1 are active (that means not greyed),
depends on the type of the currently active document window:
The brute-force analysis\index{Attack!brute-force} for DES e.~g. is only
available, if the active window is opened in the hexadecimal view. 
On the other hand the menu item ``Generate Random Numbers\dots''
is always available (even if no document is opened).

@@

%--------------------------------------------------------------------
%\nobreak
\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}
%\frame{  %TeX macht einen Rahmen (siehe viewport) drumherum -- gut zum Testen
%\includegraphics[scale=0.25, angle=270, viewport=200 30 2660 1430]
\includegraphics[scale=0.35, angle=270]
                {figures/CT1-menutree-en}
%viewport=rand-links? rand-unten breite hoehe? [bezogen auf querformat]
%}
\hypertarget{appendix-figure-menu-overview-CT1}{}
\caption{Complete overview of the menu tree of CT1 (CrypTool 1.4.31)} 
\label{appendix-figure-menu-overview-CT1}
\end{center}
\end{figure}
\clearpage

@@

\noindent Within the Startcenter you have three main ways to use the implemented functionality:
\begin{itemize}
   \item via the Wizard, which leads you to the functions.
   \item via the Workspace, where you can compose the components (e.g. an encryption method, a text input function, ...)  by yourself according to the visual programming\index{Visual programming} concept.
   \item via the template tree, which offers ready-to-run workflows.
 \end{itemize}

@@

All functions of the {\bf Algorithm Perspective} can be found in the navigation bar called ``Algorithms'' (in this perspective also at the right side).
The Algorithm Perspective contains all detail settings of the various algorithms, it especially offers post-quantum computing algorithms\index{Post-quantum computing}.

>>>>

Which of the menu items in CrypTool 1 are active (that means not greyed),
depends on the type of the currently active document window:
The brute-force analysis\index{attack!brute-force} for DES e.~g. is only
available, if the active window is opened in the hexadecimal view. 
On the other hand the menu item ``Generate Random Numbers\dots''
is always available (even if no document is opened).

@@

%--------------------------------------------------------------------
\clearpage
\begin{figure}[hb]
\begin{center}
\vspace{-30pt}

\rotatebox{90}{%
\begin{minipage}{1.57\textwidth}% So sah es für mich mittig aus.
\includegraphics[scale=0.35]{figures/CT1-menutree-en}
\hypertarget{appendix-figure-menu-overview-CT1}{}
\caption{Complete overview of the menu tree of CT1 (CrypTool 1.4.31xxxxxxxxxxxxxxxxxxx)} 
\label{appendix-figure-menu-overview-CT1}
\end{minipage}
}

\end{center}
\end{figure}
\clearpage

@@

\noindent Within the Startcenter you have three main ways to use the implemented functionality:
\begin{itemize}
   \item via the Wizard, which leads you to the functions.
   \item via the Workspace, where you can compose the components (e.g. an encryption method, a text input function, ...)  by yourself according to the visual programming\index{visual programming} concept.
   \item via the template tree, which offers ready-to-run workflows.
 \end{itemize}

@@

All functions of the {\bf Algorithm Perspective} can be found in the navigation bar called ``Algorithms'' (in this perspective also at the right side).
The Algorithm Perspective contains all detail settings of the various algorithms, it especially offers post-quantum computing algorithms\index{post-quantum computing}\index{PQC}.
